Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-01-27 paper-new/categorical-model.tex
update some definitions in the abstract model

We can specify this elegantly as a category internal to the category of CBPV models.
In particular, as in the extensional case, there is a CBPV model of ``objects''
$\mathcal M_f$ and a CBPV model of ``arrows'' $\mathcal M_{sq}$. 
There are CBPV morphisms
$r : \mathcal M_f \to \mathcal M_{sq}$ and $s, t : \mathcal M_{sq} \to \mathcal M_f$,
just as before.
%
But now, we also have a CBPV morphism $m$ from the pullback 
$\mathcal M_{sq} \times_{s = t} \mathcal M_{sq}$ to $M_{sq}$, i.e., ``composition of arrows".
In particular, this consists of a functor 
$m_{\mathcal{V}} : \vsq \times_{\sv = \tv} \vsq \to \vsq$ for composition of value
relations/squares, and a functor $m_{\mathcal{E}} : \esq \times_{\se = \te} \esq \to \esq$
for composition of computation relations/squares.
Furthermore, $s \circ m = s \circ \pi_1$ and $t \circ m = t \circ \pi_2$.

@@

\begin{definition}
  A \emph{step-0} model of intensional gradual typing consists of a category internal to the
  category of CBPV models.
\end{definition}

@@

Given the existence of the morphisms $\delta_{UB}^*$, we can define computation morphisms
$\delta_{FA}^* \in \ef(FA, FA)$ for all $A$ using the

@@

A second consequence of working intensionally is that the squares in the representable
properties must now involve a notion of ``delay" or ``perturbation'' in order to
keep the function morphisms on each side in lock-step. Intuitively, the perturbations
have no effect other than to cause the function to which they are applied to ``wait''
in a specific manner.
We formalize this notion by requiring that for each object $A$ in $\vf$,
there is a monoid of \emph{value perturbations} $\pv_A$ that is a submonoid of
$\{ f \in \vf(A,A) \mid f \bisim \id \}$.
Similarly, for each $B : \ef$ there is a monoid $\pe_B$ of
\emph{computation perturbations} that is a submonoid of 
$\{ g \in \ef(B,B) \mid g \bisim \id \}$.

@@

We require that $\delta_{FA}^* \in \pe_{FA}$ for all $A$, where $\delta_{FA}^*$ is the distinguished
morphism that is required to be present in every hom-set $\ef(FA, FA)$ per the definition
of a step-1 model.

@@

For reasons that will be made clear in the next section, perturbations must also satisfy
a property that we call the ``push-pull'' property,
which is formulated as follows. Let $c : A \rel A'$.
Given a perturbation $\delta \in \pv_A$, there is a corresponding perturbation
$\push_c(\delta) \in \pv_{A'}$. % making the following square commute:
%
Likewise, given $\delta' \in \pv_{A'}$ there is a perturbation $\pull_c(\delta') \in \pv_A$.
% making the following square commute:

@@

\begin{definition}\label{def:step-2-model}
  A \emph{step-2} model of intensional gradual typing consists of all the data of a step-1 model plus:
  \begin{enumerate}
    \item For each value type $A$, there is a monoid $\pv_A$ that is a submonoid of
    $\{ f \in \vf(A,A) \mid f \bisim \id \}$.
    \item For each computation type $B$, there is a monoid $\pe_B$ that is a submonoid of
    $\{ g \in \ef(B,B) \mid g \bisim \id \}$.
    \item For all $A$, the distinguished endomorphism $\delta_{FA}^*$ is in $\pe_{FA}$.
    % \item $\pv_A$ and a monoid homomorphism 
    %   \[ \ptbv_A : \pv_A \to \{ f \in \vf(A,A) \mid f \bisim \id \} \]
    % \item $\pe_B$ and a monoid homomorphism
    %   \[ \ptbe_B : \pe_B \to \{ g \in \ef(B,B) \mid g \bisim \id \} \]
    \item The functors $\times, \timesk$, $\arr, \tok$, $U, \Uk$, $F, \Fk$ preserve perturbations.
    \item The push-pull property holds for all $c : A \rel A'$ and all $d : B \rel B'$.
  \end{enumerate}
\end{definition}

@@

First, let $\mathcal M$ be any double category with a notion of perturbations,
i.e., for any object $X$ in $\mathcal M$ there is a monoid $P_X$ which is a
submonoid of the endomorphisms on $X$.

@@

\begin{comment}
Lastly, we require that the model satisfy a weak version of functoriality for 
the CBPV connectives $U,F,\times,\to$. 
First, we will need a definition:

\begin{definition}
  Let $c, c' : A \rel A'$. We say that $c$ and $c'$ are \emph{quasi-order-equivalent},
  written $c \qordeq c'$, if there exist perturbations $\delta^l_1, \delta^l_2 \in \pv_A$ and 
  $\delta^r_1, \delta^r_2 \in \pv_{A'}$ such that the following two squares exist:

@@

We require that the CBPV connectives $U,F,\times,\to$ are \emph{quasi-functorial} on relations,
which we specify as follows:
\begin{itemize}
  \item $U(d \comp d') \qordeq U(d)U(d')$
  \item $F(c \comp c'') \qordeq F(c)F(c')$
  \item $(cc') \to (dd') \qordeq (c \to d)(c' \to d')$
  \item $(c_1c_1') \times (c_2c_2') \qordeq (c_1 \times c_2)(c_1'\times c_2')$
\end{itemize}
\end{comment}

@@

\begin{definition}\label{def:step-3-model}
  A \emph{step-3 intensional model} consists
  of all the data of a step-2 intensional model, such that additionally:
  \begin{enumerate}
    \item There are functors $\upf : \ve \to \vf$ and $\dnf : \ee^{op} \to \ef$ % TODO: image is thin?
    \item Every value edge $c : A \rel A'$ is quasi-left-representable by $\upf(c)$ and
    every computation edge $d : B \rel B'$ is quasi-right-representable by $\dnf(d)$.
    \item \eric{Do we need the quasi-functoriality of CBPV connectives on relations?}
    % \item The CBPV connectives $U,F,\times,\to$ are quasi-functorial on relations.
  \end{enumerate}
\end{definition}

@@

\begin{definition}\label{def:step-4-model}
  % A \emph{step-$i$ intensional model with dyn} is a step-$i$ model $\mathcal M$ such that:
  A step-4 intensional model is a step-3 intensional model $\mathcal M$ such that:
  %a distinguished value object $D \in \ob(\vf)$ such that:
  %
  \begin{enumerate}
    \item There is a distinguished value object $D \in \ob(\vf)$.

    \item For each value type $A$, there is a value relation $\text{inj}_A : A \rel D$.

@@

\subsubsection{Defining an Extensional Model}\label{sec:extensional-model-definition}

>>>>

We can specify this elegantly as a category internal to the category of CBPV models
and lax morphisms, where we require that the reflexivity, source, and target morphisms
are strict.
In particular, as in the extensional case, there is a CBPV model of ``objects''
$\mathcal M_f$ and a CBPV model of ``arrows'' $\mathcal M_{sq}$. 
There are CBPV morphisms
$r : \mathcal M_f \to \mathcal M_{sq}$ and $s, t : \mathcal M_{sq} \to \mathcal M_f$,
just as before.
%
But now, we also have a CBPV morphism $m$ from the pullback 
$\mathcal M_{sq} \times_{s = t} \mathcal M_{sq}$ to $M_{sq}$, i.e., ``composition of arrows".
In particular, this consists of a functor 
$m_{\mathcal{V}} : \vsq \times_{\sv = \tv} \vsq \to \vsq$ for composition of value
relations/squares, and a functor $m_{\mathcal{E}} : \esq \times_{\se = \te} \esq \to \esq$
for composition of computation relations/squares.
Furthermore, $s \circ m = s \circ \pi_1$ and $t \circ m = t \circ \pi_2$.

As in the extensional model, we also require the existence of a natural transformation
$\mho : 1 \Rightarrow U$ such that $\mho \circ ! \ltsq{r(A)}{r(UB)} f$ for any $f : A \to UB$,
and a distinguished value type $\nat$ with morphisms $z : \mathcal V(1,\nat)$ and $s : \mathcal V(\nat,\nat)$.
% Lastly, we require the existence of a morphism $\mho_B \in \vf(\Gamma, UB)$ for all $B$
% satisfying the same ordering and commutativity properties required in the extensional model.

@@

\begin{definition}
  A \emph{step-0} model of intensional gradual typing consists of:
  \begin{itemize}
    \item A category internal to the category of CBPV models and lax morphisms,
    where we require that the morphisms $r$, $s$, and $t$ are strict.
    \item A natural transformation $\mho : 1 \Rightarrow U$ such that $\mho \circ ! \ltsq{r(A)}{r(UB)} f$ for any $f : A \to UB$.
    \item A value type $\nat$ with morphisms $z : \mathcal V(1,\nat)$ and $s : \mathcal V(\nat,\nat)$.
  \end{itemize}

@@

Given the existence of the morphisms $\delta_{UB}^*$, we can define a computation morphism
$\delta_{FA}^* \in \ef(FA, FA)$ for all $A$ by composing the unit $\eta_A \in \vf(A, UFA)$ with
the morphism $\delta_{UFA}^* \in \vf(UFA, UFA)$, and then by the adjunction
we get a computation morphism $\delta_{FA}^ \in \ef(FA, FA)$. Moreover, we get a square
$\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$ for all $c : A \rel A$.

@@

A second consequence of working intensionally is that the squares in the representable
properties must now involve a notion of ``delay" or ``perturbation'' in order to
keep the function morphisms on each side in lock-step. Intuitively, the perturbations
have no effect other than to cause the function to which they are applied to ``wait''
in a specific manner.
We formalize this notion by requiring that for each object $A$ in $\vf$,
there is a monoid of \emph{value perturbations} $P_A$ and a homomorphism of monoids
$\ptb_A : P_A \to \{ f \in \vf(A,A) \mid f \bisim \id \}$.
Similarly, for each $B : \ef$ there is a monoid $\pe_B$ of
\emph{computation perturbations} and a homomorphism of monoids 
$\ptb_B : P_B \to \{ g \in \ef(B,B) \mid g \bisim \id \}$.

@@

We will slightly abuse notation and refer to an endomorphism $f \in \vf(A, A)$ as being ``in''
the monoid of perturbations, but which we actually mean there is an element $p \in P_A$
that is mapped to $f$ under the homomorphism.

We require that $\delta_{UB}^* \in P_{UB}$ for all $A$, where $\delta_{UB}^*$ is the distinguished
morphism that is required to be present in every hom-set $\vf(UB, UB)$ per the definition
of a step-1 model.

@@

Perturbations must also satisfy a property that we call the ``push-pull'' property,
which is formulated as follows. Let $c : A \rel A'$.
Given a perturbation $\delta \in P_A$, there is a corresponding perturbation
$\push_c(\delta) \in P_{A'}$. % making the following square commute:
%
Likewise, given $\delta' \in P_{A'}$ there is a perturbation $\pull_c(\delta') \in P_A$.
% making the following square commute:

@@

\begin{definition}\label{def:step-2-model}
  A \emph{step-2} model of intensional gradual typing consists of all the data of a step-1 model plus:
  \begin{enumerate}
    \item For each value type $A$, there is a monoid $P_A$ and a homomorphism of monoids
    $\ptb_A : P_A \to \{ f \in \vf(A,A) \mid f \bisim \id \}$.
    \item For each computation type $B$, there is a monoid $\pe_B$ and a homomorphism of monoids
    $\ptb_B : P_B \to \{ g \in \ef(B,B) \mid g \bisim \id \}$.
    \item For all $B$, the distinguished endomorphism $\delta_{UB}^*$ is in $P_{UB}$.
    % \item $\pv_A$ and a monoid homomorphism 
    %   \[ \ptbv_A : \pv_A \to \{ f \in \vf(A,A) \mid f \bisim \id \} \]
    % \item $\pe_B$ and a monoid homomorphism
    %   \[ \ptbe_B : \pe_B \to \{ g \in \ef(B,B) \mid g \bisim \id \} \]
    \item The functors $\times, \timesk$, $\arr, \tok$, $U, \Uk$, $F, \Fk$ preserve perturbations.
    \item The push-pull property holds for all $c : A \rel A'$ and all $d : B \rel B'$.
  \end{enumerate}
\end{definition}

@@

First, let $\mathcal M$ be any double category with a notion of perturbations,
i.e., for any object $X$ in $\mathcal M$ there is a monoid $P_X$ with a monoid
homomorphism into the endomorphisms on $X$.

@@

Lastly, we require that the model satisfy a weak version of functoriality for 
the CBPV connectives $U,F,\times,\to$. 
First, we will need a definition:

\begin{definition}[quasi-order-equivalence]\label{def:quasi-order-equivalent}
  Let $c, c' : A \rel A'$. We say that $c$ and $c'$ are \emph{quasi-order-equivalent},
  written $c \qordeq c'$, if there exist perturbations $\delta^l_1, \delta^l_2 \in \pv_A$ and 
  $\delta^r_1, \delta^r_2 \in \pv_{A'}$ such that the following two squares exist:

@@

We require that the CBPV connectives $U,F,\times,\to$ are \emph{quasi-functorial} on relations,
which we specify as follows:
\begin{itemize}
  \item $U(d \comp d') \qordeq U(d)U(d')$
  \item $F(c \comp c'') \qordeq F(c)F(c')$
  \item $(cc') \to (dd') \qordeq (c \to d)(c' \to d')$
  \item $(c_1c_1') \times (c_2c_2') \qordeq (c_1 \times c_2)(c_1'\times c_2')$
\end{itemize}

@@

\begin{definition}\label{def:step-3-model}
  A \emph{step-3 intensional model} consists
  of all the data of a step-2 intensional model, such that additionally:
  \begin{enumerate}
    \item There are functors $\upf : \ve \to \vf$ and $\dnf : \ee^{op} \to \ef$ % TODO: image is thin?
    \item Every value edge $c : A \rel A'$ is quasi-left-representable by $\upf(c)$ and
    every computation edge $d : B \rel B'$ is quasi-right-representable by $\dnf(d)$.
    \item The CBPV connectives $U,F,\times,\to$ are quasi-functorial on relations.
  \end{enumerate}
\end{definition}

@@

\begin{definition}\label{def:step-4-model}
  % A \emph{step-$i$ intensional model with dyn} is a step-$i$ model $\mathcal M$ such that:
  A step-4 intensional model is a step-3 intensional model $\mathcal M$ such that:
  %a distinguished value object $D \in \ob(\vf)$ such that:
  %
  \begin{enumerate}
    \item There is a distinguished value object $D \in \ob(\vf)$.
    \item There are distinguished value relations 
    $\iarr{}: U(D \to F D) \rel D$ and $\inat : \nat \rel D$ and $\itimes : D \times D \rel D$
    each satisfying the retraction property up to bisimilarity.
    %  $\dnc {\injarr{}}F(\upc{\injarr{}}) \equidyn \id$.

    %\item For each value type $A$, there is a value relation $\text{inj}_A : A \rel D$.

@@

\subsubsection{Constructing an Extensional Model}\label{sec:extensional-model-definition}
